---
title: "R Notebook"
output: html_notebook
---

Install libraries if not yet previously installed.
Remove hash sign to actually run the lines.
```{r}
# install.packages("tidyverse")
# install.packages("gridExtra")
# install.packages("pheatmap")

# install.packages("BiocManager")
# BiocManager::install("flowCore")

# install.packages("devtools")
# devtools::install_github("saeyslab/FlowSOM")
# devtools::install_github("saeyslab/CytoNorm")
```

## Setup

Load the required libraries

```{r}
suppressPackageStartupMessages(library(tidyverse))
suppressPackageStartupMessages(library(gridExtra))
suppressPackageStartupMessages(library(pheatmap))
suppressPackageStartupMessages(library(flowCore))
suppressPackageStartupMessages(library(FlowSOM))
suppressPackageStartupMessages(library(CytoNorm))
```

Directory to save results to

```{r}
results <- "results_CytoNorm"
if (!dir.exists(results)) dir.create(results)
```

Set the path to the data files and extract the metadata from the filenames.
These files can be downloaded from flow repository dataset FR-FCM-Z247.

```{r}
dir <- "D:/Data/Ongoing/Stanford/data/Gates_ExtraControls"
files <- list.files(dir, pattern="Gates_P.*Control.*fcs$")
samples <- data.frame(File = files,
                      Plate = as.factor(str_match(files, "(PTLG[0-9]*)")[, 2]),
                      Stim = as.factor(str_match(files, "[0-9]_(.*)_Control")[, 2]),
                      Volunteer = as.factor(str_match(files, "_([12]).")[, 2]),
                 stringsAsFactors = FALSE)
rownames(samples) <- samples$File
samples
```

Read the marker names from the first file.  
Identify the markers of interest, ordered alphabetically, surface markers first.
```{r}
o <- capture.output(ff <- read.FCS(file.path(dir, samples$File[1])))
marker_names <- get_markers(ff, colnames(ff))

norm_ids <- c(48, 46, 43, 45, 20, 16, 21, 19, 22, 50, 47, 40, 44, 33, 17,
              11, 18, 51, 14, 23, 32, 10, 49, 27, 24, 31, 42, 37, 39, 34, 
              41, 26, 30, 28, 29, 25, 35)
norm_channels <- colnames(ff)[norm_ids]
norm_markers <- marker_names[norm_ids]

transformList <- transformList(norm_channels, cytofTransform)
transformList.reverse <- transformList(norm_channels, cytofTransform.reverse)

print(norm_markers)
```

Load the manual mapping of the data.  
See ExtractManualGating_CytoNorm.R file to generate the RDS file or download
from flowRepository.
```{r}
manual <- readRDS("D:/Data/Ongoing/Stanford/data/Gates_ExtraControls/manual.RDS")
cell_types <- c("All", levels(manual[[1]]$manual)[-1])
print(cell_types)
```

## Characterize the batch effects

### Density overview

```{r}
# Functions which use global variables, only for use in this notebook!

getDensities_plot <- function(densities, markers){
  
  densities <- lapply(densities, function(x){
    x$File <-  gsub("Norm_", "", gsub(".*/", "", x$File))
    x$Batch <- samples[x$File, "Plate"]
    x$Volunteer_Stim <- paste0(samples[x$File, "Volunteer"],
                                     "_",
                                     samples[x$File, "Stim"])
    x$Volunteer_Stim <- factor(x$Volunteer_Stim,
                               levels = c("1_Unstim", "1_IFNa_LPS",
                                          "2_Unstim", "2_IFNa_LPS"))
    x
  })
  
  plots <- list()
  for(marker in markers){
    p <- ggplot(dplyr::filter(densities$quantiles,
                              Channel == FlowSOM::get_channels(ff, marker))) +
      geom_vline(aes(xintercept = `0.05`), col = "grey", linetype="dashed") +
      geom_vline(aes(xintercept = `0.25`), col = "grey") +
      geom_vline(aes(xintercept = `0.5`), col = "#b30000", lwd = 1) +
      geom_vline(aes(xintercept = `0.75`), col = "grey") +
      geom_vline(aes(xintercept = `0.95`), col = "grey", linetype="dashed") +
      geom_line(aes(x = x, y = y),
                data = dplyr::filter(densities$densities,
                              Channel == FlowSOM::get_channels(ff, marker))) +
      facet_grid(Batch ~ Volunteer_Stim,
                 switch="y") +
      ggtitle(paste0(marker, " expression")) +
      xlab("Control and validation samples included in every plate") +
      ylab("Different plates") +
      xlim(c(-0.5, 6)) +
      theme_minimal() +
      theme(axis.text.y = element_blank(),
            strip.text.y = element_text(angle = 180))
    
    plots[[marker]] <- p
  }
  return(plots)
}
```

```{r}
res_file <- file.path(results, "densities_original.RDS")
if(!file.exists(res_file)){
  densities <- getDensities(files = file.path(dir, samples$File),
                            channels = norm_channels,
                            transformList = transformList,
                            quantileValues =  c(0.05, 0.25, 0.5, 0.75, 0.95))
  saveRDS(densities, res_file)
} else {
  densities <- readRDS(res_file)
}
```

```{r}
markers <- c("CD15", "CD66")
plots <- getDensities_plot(densities,
                           markers = c("CD15", "CD66"))

p <- do.call(gridExtra::grid.arrange, c(plots, list(nrow = 1)))

ggsave(p, 
       filename = file.path(results,
                            paste0("Density_original_",
                                   paste(markers, collapse = "_"),".pdf")),
       width = 14, height = 7)
```


```{r}
correlations <- matrix(NA, 
                       nrow = 2,
                       ncol = length(norm_markers),
                       dimnames = list(c("Unstim", "IFNa_LPS"),
                                       norm_markers))
for(channel in norm_channels){
  subset <- densities$quantiles %>% 
    dplyr::filter(Channel == channel) 
  sample_types <- str_match(subset$File, "PTLG[0-9]*_(.*).fcs")[,2]
  medians <- subset$`0.5`
  correlations["Unstim", marker_names[channel]] <- cor(medians[sample_types == "Unstim_Control_1"],
                                                       medians[sample_types == "Unstim_Control_2"])
  correlations["IFNa_LPS", marker_names[channel]] <- cor(medians[sample_types == "IFNa_LPS_Control_1"],
                                                           medians[sample_types == "IFNa_LPS_Control_2"])
}

apply(correlations, 1, mean)
apply(correlations, 1, sd)
```

### Cell type specific densities

```{r fig.width = 14, fig.height = 7 }
cell_types_to_plot <- c("B cells", "Monocytes")
marker <- "HLADR"

res_file <- file.path(results, 
                          paste0("densities_original_",
                                 paste(cell_types_to_plot, collapse = "_"),
                                 ".RDS"))
if (!file.exists(res_file)) {
  
  densities_selected <- list()
  
  for (cell_type in cell_types_to_plot) {
    cell_type_selection <- lapply(manual, function(x){
      x$matrix[, cell_type]
    })
    
    names(cell_type_selection) <- file.path(dir, names(cell_type_selection))
    
    densities_selected[[cell_type]] <- 
      getDensities(files = file.path(dir, samples$File),
                   channels = norm_channels,
                   transformList = transformList,
                   quantileValues =  c(0.05, 0.25, 0.5, 0.75, 0.95),
                   selection = cell_type_selection)
  }

  saveRDS(densities_selected, res_file)
} else {
  densities_selected <- readRDS(res_file)
}
```

```{r}
plots <- list()
for(cell_type in cell_types_to_plot){
  p <- getDensities_plot(densities_selected[[cell_type]],
                         marker)[[1]] + 
       ggtitle(paste0(marker, " expression in ", cell_type)) +
       xlim(c(2.5, 6))
  
  plots[[cell_type]] <- p
}

p <- do.call(gridExtra::grid.arrange, c(plots, list(nrow = 1)))

ggsave(p, 
       filename = file.path(results,
                            paste0("Density_original_",
                                   marker, "_",
                                   paste(cell_types_to_plot, collapse = "_"),
                                   ".pdf")),
       width = 14, height = 7)
              
```


BC densities for reviewer
```{r}
BC_channels <- colnames(ff)[grep("BC", get_markers(ff, colnames(ff)))]
densities_BC <- getDensities(files = file.path(dir, samples$File),
                             channels = BC_channels,
                             transformList = transformList(BC_channels, 
                                                           cytofTransform),
                             quantileValues =  c(0.05, 0.25, 0.5, 0.75, 0.95))

plots <- getDensities_plot(densities_BC,
                           markers = get_markers(ff, BC_channels))

p <- do.call(gridExtra::grid.arrange, c(plots, list(nrow = 2)))

ggsave(p, 
       filename = file.path(results,
                            paste0("Density_original_",
                                   paste(get_markers(ff, BC_channels), 
                                         collapse = "_"),
                                   ".pdf")),
       width = 21, height = 14)

```