---
title: "R Notebook"
output: html_notebook
---

This notebook assumes you have run the setup part of
1_data_exploration_of_control_and_validation_samples.Rmd first,
as it makes use of variables created in that notebook.


## Evaluation of the FlowSOM clustering

```{r}
train_files <- samples %>% 
  dplyr::filter(Volunteer == 1) %>% 
  dplyr::filter(Stim == "Unstim") %>%
  dplyr::pull(File)
```

```{r}
res_file <- file.path(results, "prepareFlowSOM.RDS")
if (!file.exists(res_file)) {
  fsom <- prepareFlowSOM(file.path(dir, train_files),
                         norm_channels,
                         nCells = 1000000,
                         FlowSOM.params = list(xdim = 15,
                                               ydim = 15,
                                               nClus = 25,
                                               scale = FALSE),
                         transformList = transformList,
                         plot = TRUE,
                         seed = 1)
  saveRDS(fsom, file.path(results, "prepareFlowSOM.RDS"))
} else {
  fsom <- readRDS(res_file)
}
```

```{r}
# Test different amounts of metaclusters
pdf(file.path(results, "CV_test.pdf"),
    width = 11, height = 7)
cv_test <- testCV(fsom,
                  cluster_values = 3:50,
                  plot = TRUE,
                  verbose = TRUE)
dev.off()

cv_test$cvs$`25`
```

## Run normalization on the samples

### CytoNorm normalization

Experiment setup

```{r}
FlowSOM.params <- list(nCells = 1000000,
                       xdim = 15,
                       ydim = 15,
                       nClus = 25,
                       scale = FALSE)

norm.params <- list("QuantileNormLimited" = list(nQ = 101,
                                                 limit = c(0,8),
                                                 goal = "PTLG021"),
                    
                    "QuantileNorm" = list(nQ = 101,
                                          goal = "PTLG021"),
                    
                    "MinMaxNorm" = list(nQ = 2,
                                        quantile_values = c(0, 1),
                                        goal = "PTLG021"),

                    "QMinMaxNorm" = list(nQ = 2,
                                         quantile_values = c(0.001, 0.999),
                                         goal = "PTLG021"))

exp_setups <- expand.grid(clustering = c("Cluster", ""),
                          norm = names(norm.params),
                          stim = c("Unstim_and_IFNa_LPS"),
                          volunteer = c(1,2))

rownames(exp_setups) <- gsub("^_", "", 
                             apply(exp_setups, 1, paste, collapse = "_"))
exp_setups
```

Setup for testing control samples with different ranges
```{r}
FlowSOM.params <- list(nCells = 1000000,
                       xdim = 15,
                       ydim = 15,
                       nClus = 25,
                       scale = FALSE)

norm.params <- list("QuantileNormLimited" = list(nQ = 101,
                                                 limit = c(0,8),
                                                 goal = "PTLG021"))

exp_setups <- expand.grid(clustering = c("Cluster"),
                          norm = names(norm.params),
                          stim = c("Unstim_and_IFNa_LPS", "Unstim", "IFNa_LPS"),
                          volunteer = c(1,2))

rownames(exp_setups) <- gsub("^_", "", 
                             apply(exp_setups, 1, paste, collapse = "_"))

exp_setups
```

Training 

```{r}
if(!dir.exists(file.path(results, "models")))
  dir.create(file.path(results, "models"))

if(file.exists(file.path(results, "models", "times.rds"))){
  times <- readRDS(file.path(results, "models", "times.rds"))
} else {
  times <- list()
}

for(exp in rownames(exp_setups)){
  if(!dir.exists(file.path(results, "models", exp))){
    dir.create(file.path(results, "models", exp))
  }
  print(paste0("Training ", exp))
  train_data <- samples %>% 
    dplyr::filter(Volunteer == exp_setups[exp, "volunteer"]) %>% 
    dplyr::filter(grepl(gsub("_and_", "|",exp_setups[exp, "stim"]), Stim))
  
  if(exp_setups[exp, "clustering"] == "Cluster"){
    
    t <- system.time(capture.output(
      model <- CytoNorm.train(files = file.path(dir, 
                                                train_data$File),
                              labels = train_data$Plate,
                              channels = norm_channels,
                              transformList = transformList,
                              FlowSOM.params = FlowSOM.params,
                              normMethod.train = QuantileNorm.train,
                              normParams = norm.params[[exp_setups[exp, "norm"]]],
                              seed = 1,
                              outputDir = file.path(results, "models", exp),
                              plot = TRUE)))
  } else {
    t <- system.time(capture.output(
      model <- do.call(QuantileNorm.train,
                       c(list(files = file.path(dir, 
                                                train_data$File),
                              labels = train_data$Plate,
                              channels = norm_channels,
                              transformList = transformList,
                              plot = TRUE),
                         norm.params[[exp_setups[exp, "norm"]]]))))
    
  }
  
  print(t)
  times[[exp]] <- t
  saveRDS(times,
          file = file.path(results,
                           "models",
                           "times.rds"))
  saveRDS(model,
          file = file.path(results,
                           "models",
                           exp,
                           "GatesExtraControls_model.rds"))
  
}
```

Normalization

```{r}
dir.create(file.path(results, "data"))

for(exp in rownames(exp_setups)){
  dir.create(file.path(results, "data", exp))
  print(paste0("Normalizing ", exp))
  model <- readRDS(file.path(results,
                             "models",
                             exp,
                             "GatesExtraControls_model.rds"))
  test_data <- samples %>% 
    dplyr::filter(Volunteer != exp_setups[exp, "volunteer"])
  
  if(exp_setups[exp, "clustering"] == "Cluster"){
    
    t <- system.time(capture.output(
      CytoNorm.normalize(model = model,
                         files = file.path(dir, test_data$File),
                         labels = test_data$Plate,
                         transformList = transformList,
                         transformList.reverse = transformList.reverse,
                         outputDir = file.path(results, "data", exp),
                         normMethod.normalize = QuantileNorm.normalize)))
  } else { 
    
    t <- system.time(capture.output(
      QuantileNorm.normalize(model = model,
                             files = file.path(dir, test_data$File),
                             labels = test_data$Plate,
                             transformList = transformList,
                             transformList.reverse = transformList.reverse,
                             outputDir = file.path(results, "data", exp))))
    
  }
  print(t)
}
```

```{r}
emds <- list()

for(volunteer in c(1,2)){
  for(stim in c("Unstim", "IFNa_LPS")){
    print(paste0("Evaluating original"))
    toEvaluate <- samples %>% 
      dplyr::filter(Volunteer != volunteer) %>% 
      dplyr::filter(Stim == stim)
    
    o <- capture.output(
      emds[[paste0("Original_", volunteer, "_", stim)]] <- 
        emdEvaluation(file.path(dir,
                                toEvaluate$File),
                      transformList = transformList,
                      manual = lapply(manual, function(x){x$manual}),
                      channels = norm_channels))
  }
}

for (exp in rownames(exp_setups)) {
  for(stim in c("Unstim", "IFNa_LPS")){
    print(paste0("Evaluating ", exp))
    toEvaluate <- samples %>% 
      dplyr::filter(Volunteer != exp_setups[exp, "volunteer"]) %>% 
      dplyr::filter(Stim == stim)
    
    o <- capture.output(
      emds[[paste0(exp, "_", stim)]] <- emdEvaluation(file.path(results, "data", exp, 
                                             paste0("Norm_",
                                                    toEvaluate$File)),
                                   transformList = transformList,
                                   manual = lapply(manual, function(x){x$manual}),
                                   channels = norm_channels))
  }
}

saveRDS(emds,
        file = file.path(results, paste0("emds_diff_methods.Rds")))
```

```{r}
emds <- readRDS(file.path(results,
                         "emds_diff_methods.Rds"))
emds_g <- lapply(names(emds), function(state){
  col_name <- paste0("EMD_",state)
  gather(data.frame(Population = rownames(emds[[state]]), 
                    emds[[state]])[-1,], 
         "Channel", col_name, -"Population")  %>% 
    magrittr::set_colnames(c("Population", "Channel", col_name))
}) %>% magrittr::set_names(names(emds))

emds_g <- cbind(emds_g[[1]][,c(1,2)],
                do.call(cbind, lapply(emds_g, function(x) x[,3])))
rownames(emds_g) <- paste0(emds_g[,1], "_", get_markers(ff, emds_g[,2]))


reduction <- list()
for(exp in colnames(emds_g)[-c(1,2)]){
  exp_original <- gsub(".*_([12].*)", "Original_\\1", exp)
  affected <- (emds_g[,exp] > 2) | (emds_g[ ,exp_original] > 2)
  reduction[[exp]] <- data.frame(population_marker = rownames(emds_g)[affected],
                                 reduction = ((emds_g[,exp_original] - emds_g[,exp]) /
                                                emds_g[,exp_original])[affected],
                                 exp = exp,
                                 method = gsub("_[12].*", "", exp),
                                 subset = gsub("^.*_([12].*)", "\\1", exp))
}
reduction <- do.call(rbind, reduction)

reduction_overall <- cbind("mean" = tapply(reduction$reduction, 
                                         reduction$method, 
                                         mean),
                           "sd" = tapply(reduction$reduction,
                                       reduction$method,
                                       sd))
for (volunteer in c(1,2)) {
  for (stim in c("Unstim", "IFNa_LPS")) {
    for(f in c("mean", "sd")){
      subset <- reduction %>% 
                 dplyr::filter(subset == paste0(volunteer, "_", stim)) 
      reduction_subset <- tapply(subset$reduction, 
                                 subset$method, 
                                 eval(parse(text = f)))
     
      
      reduction_overall <- cbind(reduction_overall,
                                 matrix(reduction_subset,
                                        dimnames = list(names(reduction_subset),
                                                        paste0(volunteer, "_", stim, "_", f))))
        
    }

  }
}
                           
print(reduction_overall)
paste0(round(reduction_overall[, grep("mean",colnames(reduction_overall))], 2),
       " (+- ", 
       round(reduction_overall[, grep("sd",colnames(reduction_overall))], 2),
       ")")
pheatmap::pheatmap(reduction_overall[-1, grep("_mean", colnames(reduction_overall))],
                   cluster_rows = FALSE,
                   cluster_cols = FALSE,
                   display_numbers = TRUE,
                   number_color = "white")
```


```{r}
for(e in unique(reduction$exp)){
 subset <- reduction %>% 
   dplyr::filter(exp == e)
 plot(density(subset$reduction), main = e)
}
```
```{r}
for (volunteer in c(1,2)) {
  for (stim in c("Unstim", "IFNa_LPS")) {
      subset <- reduction %>% 
                 dplyr::filter(subset == paste0(volunteer, "_", stim)) 
      reduction_subset <- tapply(subset$reduction, 
                                 subset$method, 
                                 eval(parse(text = f)))
     
      
      reduction_overall <- cbind(reduction_overall,
                                 matrix(reduction_subset,
                                        dimnames = list(names(reduction_subset),
                                                        paste0(volunteer, "_", stim, "_", f))))
        
  }
  
}

  
```

```{r}
plot_emds <- function(exp_1, exp_2, emds_g, title = NULL, subtitle = NULL){
  emds_g$affected <- rep("Positive", nrow(emds_g))
  emds_g$affected[emds_g[,exp_1] > emds_g[,exp_2]] <- "Negative"
  emds_g$affected[emds_g[,exp_1] < 2 & emds_g[,exp_2] < 2] <- "No"
  
  if(is.null(title)){
    method <- stringr::str_match(exp_1, "(.*)_([1-2].*)")[,2]
    subset <- stringr::str_match(exp_1, "(.*)_([1-2].*)")[,3]
    title <- "Average change: "
    subtitle <- paste0(round(reduction_overall[method, paste0(subset,"_mean")], 4), 
                       " (±", round(reduction_overall[method, paste0(subset,"_sd")],4), ")")
  }
  
  p <- ggplot(emds_g) +
    geom_point(aes_string(x = exp_1, 
                          y = exp_2, 
                          col = "affected")) +
    geom_rect(data = data.frame(xmin = 0, xmax = 2, ymin = 0, ymax = 2),
              aes(xmin = xmin, xmax = xmax, ymin = ymin, ymax = ymax),
              fill = "#555555", alpha = 0.3) +
    scale_x_continuous(limits = c(0, 15)) +
    scale_y_continuous(limits = c(0, 15)) +
    scale_color_manual(values = c("No" = "grey", 
                                  "Positive" = "black", 
                                  "Negative" = "#b30000")) +
    ggplot2::geom_abline(lwd = 1.5) +
    coord_fixed() +
    ggtitle(title, subtitle) +
    theme_minimal() +
    theme(plot.subtitle = element_text(hjust = 0.5),
          legend.position = "none") 
  emds_g$affected <- NULL
  return(p)
}
```

```{r}

plots <- list()
for (test in c("Unstim", "IFNa_LPS")) {
  for (model in c("Unstim_and_IFNa_LPS", "Unstim", "IFNa_LPS")){
    plots[[paste0(model,"_",test)]] <- plot_emds(paste0("Cluster_QuantileNormLimited_", model, "_1_", test),
                                                 "Original_1_Unstim",
                                                 emds_g) +
      xlab("EMD after normalization") +
      ylab("EMD before normalization") +
      ggtitle(paste0("Trained on ", gsub("_and_", " and ", model)), 
              subtitle = paste0("Validated on ", test)) + 
      theme(plot.title = element_text(size = 11),
            plot.subtitle = element_text(size = 11))
  }
}

p <- do.call(grid.arrange, c(plots, list(nrow = 2)))
ggsave(file.path(results,
                 "EMD_Unstim_vs_IFNa_LPS.pdf"),
       p,
       width = 9, height = 6)
p
```

```{r}
plot_titles <- c(c("Cluster_QuantileNormLimited" = "Cluster + 101 quantiles",
                   "QuantileNormLimited" = "101 quantiles",
                   "Cluster_QMinMaxNorm" = "Cluster + 2 quantiles",
                   "QMinMaxNorm" = "2 quantiles"))

plots <- list()
for (volunteer in c(1,2)){
  for (test in c("Unstim", "IFNa_LPS")) {
    for(method in c("Cluster_QuantileNormLimited", "QuantileNormLimited",
                    "Cluster_QMinMaxNorm", "QMinMaxNorm")){
      plots[[paste0(method,"_",volunteer,"_",test)]] <- plot_emds(paste0(method, "_Unstim_and_IFNa_LPS_", volunteer, "_", test),
                                                   paste0("Original_",volunteer,"_", test),
                                                   emds_g) +
        xlab("EMD after normalization") +
        ylab("EMD before normalization") +
        ggtitle(plot_titles[method], 
                subtitle = paste0("Validated on donor ", volunteer, " ", test)) + 
        theme(plot.title = element_text(size = 11),
              plot.subtitle = element_text(size = 11))
    }
  }
}
p <- do.call(grid.arrange, c(plots, list(nrow = 4)))
ggsave(file.path(results,
                 "EMD_methods.pdf"),
       p,
       width = 12, height = 12)
```